---
title: "Designing and evaluating a spatial multi-omics workflow"
author: "Tianyao Lu"
output:
  BiocStyle::html_document:
    toc: true
vignette: >
  %\VignetteIndexEntry{Designing and evaluating a spatial multi-omics workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "##", fig.width = 9, fig.height = 5)
library(SpaMTP)
```

# Analysis questions determine the workflow

A comprehensive report should expose the consequences of analytical choices.
Its purpose is not to produce an attractive embedding or maximize the number of
significant features. The workflow follows the analysis chain of the historical
mouse-brain vignette: inspect measurements and correspondence; compare individual,
spatial and integrated representations; characterize regions; examine paired
features; then connect measured identities to pathways and networks.

| Question | Native interface | Evidence to inspect |
|:--|:--|:--|
| What was measured and retained? | `normalizeSMData()`, native containers and image access | Selected layer, QC exclusions, scale and acquisition provenance |
| What changes when spatial information is added? | `runMetabolicPCA()`, `scaleSMData()`, `runSpatialGraphPCA()` | Same selected features and all retained observations in both representations |
| Does integration improve the representation? | `multiOmicIntegration()`, `getKmeanClusters()` | Individual versus joint structure; external-reference ARI, approximate silhouettes and sensitivity across K |
| Which features characterize a region? | `findAllDEMs(method="markers")`, `demsHeatmap()` | Pairwise AUC, Cohen effects, detection, mean expression and spatial-block sensitivity |
| Which differences reproduce across biological samples? | `findAllDEMs(method="replicate")` | Explicit design, replicate means, effect intervals and the complete BH family |
| Is an association explained by regional composition? | `findCorrelatedFeatures()`, `plot3DFeature()` | Raw versus covariate-residual correlations and matched spatial layers |
| Which members support a pathway? | `buildPathwayIndex()`, `createPathwayAssay()`, `findRegionalPathways()` | Measured coverage, ambiguity exclusions, complete ranks and leading edges |
| Are pathway members coordinated and where are they? | `runRAMPGeseca()`, `plotRegionalPathways()`, `pathwayNetworkPlots()` | The same identity assay and membership index in every downstream view |

The evaluated examples below are **synthetic teaching data**, not biological
findings. The installed `workflows/prepare_mouse_brain.R` and
`workflows/run_mouse_brain.R` scripts reconstruct real FMP10/Visium and DHB/Visium
analyses from registered SpaMTPData and SpaMTPdb resources.

# Supply modalities and study design explicitly

```{r inputs}
x <- SpaMTPData::spaMTPExampleData(paired = TRUE)
protein <- rbind(marker1 = c(1:6, 8:13), marker2 = c(6:1, 2:7),
                 marker3 = rep(c(2, 7, 4), 4))
colnames(protein) <- colnames(x)
SingleCellExperiment::altExp(x, "protein") <-
    SingleCellExperiment::SingleCellExperiment(assays = list(counts = protein))
spec <- list(
    main = list(type = "metabolomics", layer = "counts"),
    transcriptome = list(type = "transcriptomics", layer = "counts"),
    protein = list(type = "proteomics", layer = "counts")
)
```

The workflow preserves the supplied layers. Its declared baseline uses native
library-size normalization followed by conversion to log2 units for raw MS/RNA
counts, and log2(1 + value) for protein counts. For already transformed data,
choose `normalization="none"`. This baseline does not replace vendor-specific
spectral processing, RNA count QC or a platform-appropriate normalization design.
In particular, logged values must not be logged again or labelled raw counts.

Biological replication is an explicit metadata field. A pixel, technical pool,
spatial block, image channel or integration cluster does not create an independent
specimen. Transcriptomics, metabolomics and proteomics can be declared as three
modalities; H&E, DAPI and RGB channels remain contextual images unless their
measurements support a separate stated assay model.

# Compare representations using the same observations and features

```{r workflow}
fit <- runSpaMTPWorkflow(x, spec, regions = "region", clusters = 2,
    npcs = 2, association_features = 3,
    coordinate_units = "synthetic grid units",
    structure = list(primary = "spatial", k_grid = c(2, 3),
                     reference = "region", umap = FALSE,
                     stability_blocks = 3))
fit$structure$metrics
```

`runMetabolicPCA(features=...)` now records all selected features explicitly.
Ordinary and graph PCA use the same variable features, centred and scaled in the
same way. Graph PCA uses a sparse Laplacian for every retained observation and
keeps different samples in separate spatial graphs. It is not fitted on a small
report preview and then discarded: graph embeddings feed the spatial joint
representation and can be selected as the downstream primary representation.

The joint representation uses `multiOmicIntegration()` to concatenate standardized
modality embeddings. This native implementation is **not Seurat WNN**. Recovering
and benchmarking WNN requires a separate, validated method implementation.
UMAP is a visualization of a representation; clusters and metrics use its complete
retained dimensions. A nicer-looking UMAP is not a performance test.

The number of clusters and primary method are chosen before examining reference
ARI. Approximate silhouettes assess separation, while ARI compares to supplied
external annotation. Spatial-neighbor agreement is a smoothness diagnostic; spatial
regularization increases it by construction. Spatial-block omission refits
clustering **conditional on full-data embeddings**. It does not validate feature
selection, normalization or generalization to new specimens.

# Region characterization is richer than a mean difference

```{r markers}
markers <- fit$region_analysis$modalities$main$markers
head(markers$DEMs[, c("gene", "cluster", "logFC", "mean_auc",
                     "min_auc", "mean_cohen", "direction_stability")])
demsHeatmap(markers, n = 2, only.pos = TRUE, logfc.threshold = 0,
    order.by = "mean_auc", fontsize_row = 9, fontsize_col = 10)
```

The native marker mode delegates pairwise scoring to `scran::scoreMarkers()`.
AUC measures pairwise concordance, including half credit for ties. Mean and minimum
AUC distinguish broadly specific markers from features that separate only some
regions. Cohen effects account for within-region variability. Summaries equally
weight other eligible regions so that the largest anatomical compartment does not
automatically dominate the reference. Detection fractions refer to non-zero values
on the explicitly selected layer; residualized layers need different interpretation.

Coordinate-only blocks support a leave-one-block-out sensitivity analysis. The
result records the range of Cohen effects, marker-rank changes, direction agreement
and omitted fits that were ineligible because a region became too small. These
ranges are **not confidence intervals**. Region labels are held fixed, including
when learned from these data, and no pixel-based P values are manufactured.

`demsHeatmap()` uses the analysis expression matrix directly. Its FDR ordering now
sorts numerical FDR; it no longer sorts a regulation label or applies CPM to an
already transformed marker matrix. The heatmap, detection plot, spatial feature
maps and interactive explorer use the same ranked feature identifiers.

# Keep biological-replicate inference separate

```{r replicate-design, eval=FALSE}
# Use only when the metadata identifies independent specimens and conditions.
replicate_de <- findAllDEMs(real_data, ident = "condition", assay = "main",
    slot = "logcounts", method = "replicate", replicate = "patient_id",
    contrasts = list(treatment_control = list(
        numerator = "treated", denominator = "control", paired = TRUE)))
```

This mode fits limma to equally weighted biological-replicate means of the selected
values. Paired contrasts retain complete pairs; at least three complete pairs (or
three independent replicates per arm) are required. The result retains the design,
coefficient, residual degrees of freedom, replicate expression matrix, pointwise
moderated intervals, and every tested feature. BH correction is within modality
and contrast. A `logFC` column is a difference on the selected workflow scale;
it is not automatically a fold change in raw abundance. RNA count models may
instead require independently designed count pseudobulking and edgeR/DESeq2.

The historical random-pool mode remains available for compatibility and emits a
warning about technical partitions. It is never used by the automatic workflow.

# Ask whether colocalization survives regional composition

```{r associations}
head(fit$associations[[1]])
```

`findCorrelatedFeatures(covariates=...)` retains raw Pearson correlation and the
correlation between both features after regression on the declared sample,
replicate and region fields. The workflow queries regional marker anchors against
all PCA-selected target features. It records those selection sets and the matched
observations. Strong raw association can disappear after removing a common regional
shift. A residual association can still reflect spatial autocorrelation or other
unmeasured structure, and selected pairs require independent validation.

Exact feature identifiers work for RNA and protein as well as metabolites; no
artificial m/z metadata is needed. The report includes paired scatter plots and
native `plot3DFeature()` views of the same selected pairs.

# Map identities before pathway statistics

```{r pathway-chain, eval=FALSE}
# Pin a compatible organism-specific index and database resources.
identity <- createPathwayAssay(real_data, analyte_type = "genes",
    assay = "RNA", slot = "logcounts", new_assay = "gene_ids",
    pathway_index = index, organism = "Homo sapiens")
identity_markers <- findAllDEMs(identity, ident = "region", assay = "gene_ids",
    slot = "logcounts", method = "markers")
# Build a feature-by-region matrix of the complete identity-level effects.
ranks <- identity_markers$expression
for (region in colnames(ranks)) {
    tab <- identity_markers$DEMs[identity_markers$DEMs$cluster == region, ]
    ranks[, region] <- tab$logFC[match(rownames(ranks), tab$gene)]
}
regional <- findRegionalPathways(identity, ident = "region",
    ranks = list(genes = ranks), pathway_index = index)
plotRegionalPathways(regional, database = database)
```

The ranks interface uses the complete measured feature ranking, with no FDR or
foreground prefilter. It retains every used rank and the complete eligible pathway
family. Competitive feature-set P values do not establish reproducibility across
specimens. Joint gene/metabolite ranks can be RMS-scaled per modality, but pathways
with many measured genes can still be dominated by transcriptomic coverage.

For metabolomics, `createPathwayAssay(metabolite_ambiguity="exclude")` excludes
mass features with multiple candidate compound identities. Original candidates,
exclusions, aggregation weights and measured members remain inspectable. The
workflow uses this policy by default; the public native interface retains the
historical `expand` option. Sparse matrices stay sparse during identity aggregation.
The network viewer can consume this identity assay directly, preventing re-expansion
of excluded mass candidates. GESECA, scores and networks therefore use the same
identity definition as regional enrichment.

The pathway dot plot uses **adjusted** P values for significance outlines and
`1 - Jaccard similarity` for pathway ordering. It handles a single pathway. Mouse
genes are not silently uppercased into a human HGNC index; the registered mouse
recipes currently show compound pathways until an explicit compatible gene identity
or orthology reference is supplied.

# Run the real mouse-brain reference recipes

```{r mouse-recipes, eval=FALSE}
prepare <- system.file("workflows", "prepare_mouse_brain.R", package = "SpaMTP")
run <- system.file("workflows", "run_mouse_brain.R", package = "SpaMTP")
# Rscript prepare NATIVE_1.1.0_DIR NEW_INPUT_DIR [SPATIAL_CONFIG.json]
# Rscript run NEW_INPUT_DIR SPAMTPDB_3.0.7_DIR NEW_OUTPUT_DIR fmp10
# Rscript run NEW_INPUT_DIR SPAMTPDB_3.0.7_DIR NEW_OUTPUT_DIR dhb
```

Preparation reads only the pinned native SpaMTPData 1.1.0 resources, offline.
It does not load Seurat or deserialize historical Seurat objects. H&E is optional:
the published native resources contain expression and coordinates but omit
optical rasters. Independent image files can be attached with `addSpatialImage()`;
their checksums, explicit scale factors and source provenance are retained.

For FMP10/Visium pixel-overlap mapping, supply a measured Visium spot radius in
the original image coordinate system. A nearest-neighbour distance is not a
spot diameter. Without a radius, preparation still produces usable DHB inputs,
but the FMP10 runner stops with an explicit geometry requirement. The optional
JSON configuration accepts `source`, `target_radius` and `images`. Image paths
are resolved relative to the R process working directory, not the JSON file.

```{r mouse-spatial-config, eval=FALSE}
# Set these values from the original acquisition metadata, not fitted to a plot.
config <- list(
    source = "Accession and acquisition metadata used for these image scales",
    target_radius = spotDiameterFullres / 2,
    images = list(mouse_brain_visium = list(
        path = "tissue_lowres_image.png", scale_factor = lowresScaleFactor))
)
jsonlite::write_json(config, "spatial-config.json", auto_unbox = TRUE,
                    pretty = TRUE, digits = NA)
# source(prepare) defines prepareMouseBrain() without running it.
source(prepare)
prepareMouseBrain("native-resources", "new-inputs", "spatial-config.json")
```

Pairing is documented for these registered source specimens; arbitrary datasets
are not paired merely because both are mouse brains. DHB already contains paired
modalities. Both runners re-annotate with the pinned chemical database and an
explicit positive-mode DHB or FMP-10 index, rather than using the archived
mass-only annotation cache. Ambiguous compound assignments remain excluded;
mouse RNA is retained without human-HGNC or implicit orthology conversion.

The installed network-only demo is another native entry point:

```{r mouse-network-demo, eval=FALSE}
demo <- system.file("examples", "pathway_network_mouse_brain_demo.R",
                    package = "SpaMTP")
source(demo)
networks <- runMouseBrainNetworkDemo("native-resources", "database-resources",
                                    "new-network-output", ident = "RegionLoupe")
```

It uses `findAllDEMs(method="markers")` for descriptive regional effects,
complete compound rankings for competitive pathway enrichment, and
`pathwayNetworkPlots()` for separate leading-edge and annotation-coverage views.
It does not assign biological-replicate significance to individual pixels or
map mouse gene symbols to human genes. Outputs include the native analysis RDS,
annotation/mapping audits and, when pathways are eligible, offline HTML networks.
No eligible pathways is reported explicitly rather than relaxing the filters.

# Render and retain a reviewable analysis

```{r report, eval=FALSE}
report <- renderSpaMTPReport(fit, "new_report_directory")
```

Reports follow the analytical sequence above. They include a linked region/DE
explorer with feature search, directional AUC/effect/stability/FDR filters, spatial
zoom and pan, coordinated feature profiles and complete CSV exports. Native 3D and
network views are available when their input requirements are satisfied. Figures
are embedded as PNGs and also exported as PDFs. Open the HTML in a browser or an IDE
HTML preview that permits JavaScript; Markdown previews may disable interaction.
The output directory also contains the complete workflow RDS, settings, inputs used
by each stage, source and database provenance, and file checksums.

For methods, see the [Bioconductor marker-scoring documentation](https://bioconductor.org/books/release/OSCA.basic/marker-detection.html)
and the [GraphPCA paper](https://doi.org/10.1186/s13059-024-03429-x).
