---
title: "Auditable gene mapping with SpaMTPdb and SpaMTPData"
output:
  BiocStyle::html_document:
    toc: true
vignette: >
  %\VignetteIndexEntry{Auditable gene mapping with SpaMTPdb and SpaMTPData}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "##")
```

## Reference ownership and reproducibility

SpaMTPdb provides RaMP annotation tables and a separate versioned HGNC gene
reference. SpaMTPData supplies native experiment data and declared species.
SpaMTP resolves identifiers, annotates row metadata and runs pathway analyses.
Use SpaMTP >= 0.99.6, SpaMTPdb >= 0.99.4 and SpaMTPData >= 0.99.5 together.

```{r registry}
library(SpaMTP)
SpaMTPdb::spaMTPdbGeneReference(metadata = TRUE)
SpaMTPdb::spaMTPdbVersion()
```

The HGNC reference uses a fixed quarterly archive, with its own version and
checksum. It does not change the existing RaMP 3.0.7 or experiment 1.1.0 files.
HGNC mapping is for **Homo sapiens**; it does not convert mouse genes to human
orthologues. Native experiments loaded by SpaMTPData carry their registry species
in `metadata(x)$SpaMTPData`, including their nested experiments. Species
checks therefore still apply after extracting RNA with `altExp()`.

## Build once, inspect before testing

The following resource-loading examples are not evaluated while building this
vignette. Local files and verified caches can be used with `offline = TRUE`.

```{r reference, eval=FALSE}
database <- loadSpaMTPDatabase(
  c("source_df", "analyte", "analytehaspathway", "pathway"),
  version = "3.0.7", local_dir = "/path/to/RaMP/3.0.7", offline = TRUE
)
reference <- SpaMTPdb::spaMTPdbGeneReference(
  version = "2026.7.7", local_dir = "/path/to/gene-reference", offline = TRUE
)
index <- buildGeneMappingIndex(database$source_df, gene_reference = reference)
mapping <- mapGeneIdentifiers(
  c("gene_symbol:AATK", "entrez:9625", "gene_symbol:BANK1"), index = index
)
mapping
```

The index records a fingerprint of the source table's gene identifiers and
common names. Pathway workflows reject a cached index built from different
identity assignments, even if the RaMP ID set is unchanged. Rebuild older
saved indices with `buildGeneMappingIndex()`; changing row order or duplicating
identical source rows does not invalidate the fingerprint.

Symbols, previous symbols and aliases are resolved to an HGNC gene. An approved
symbol takes priority over other genes' aliases. Entrez, HGNC, UniProt and
versioned Ensembl gene IDs are supported. A shared protein ID can be resolved
by a compatible gene-specific ID in the same RaMP record. Conflicting specific
IDs quarantine the record rather than joining different genes.

`gene_id` is the stable HGNC identity. `ramp_ids` lists its non-conflicting RaMP
records, and `canonical_ramp_id` is one representative used to count the gene
once. For example, AATK's UniProt and Entrez records in RaMP 3.0.7 belong to one
HGNC gene, so their pathway memberships are united before enrichment.

Inspect `status`, `method`, `candidate_gene_ids` and `candidate_ramp_ids`.
The latter exposes raw RaMP candidates for direct IDs and crossreferences.
A shared crossreference with unresolved or conflicting competitors remains
ambiguous even when only one candidate has a known HGNC identity.
`not_in_ramp` means HGNC
recognizes the gene but no usable RaMP record was found; `ambiguous` and
`conflicting_ramp_records` require review. `mapped_ramp_only` denotes a raw
RaMP record whose HGNC identity was not resolved. A warning is not evidence
that the gene has no biological function or pathway membership.

## Annotate experiment features without altering expression

```{r experiment, eval=FALSE}
spe <- SpaMTPData::spaMTPData("human_brain_integrated", version = "1.1.0")
SingleCellExperiment::altExpNames(spe)
# Select the actual transcriptome altExp shown above. If identifiers live in
# a rowData column, set id_column; otherwise row names are used.
spe <- annotateGeneIdentifiers(
  spe, assay = "SPT", index = index, organism = "Homo sapiens"
)
rna <- SingleCellExperiment::altExp(spe, "SPT")
SummarizedExperiment::rowData(rna)
S4Vectors::metadata(rna)$gene_mapping
```

Mapping columns remain aligned with original features. Counts, feature names,
coordinates and paired modalities are preserved. If multiple expression rows
map to the same gene, annotation records that fact without summing expression.
Resolve those features before differential analysis. Regional pathway and
network workflows reject different differential statistics for the same gene
within a cluster rather than selecting an arbitrary row.

## Use the whole measured panel as the universe

```{r enrichment, eval=FALSE}
result <- fishersPathwayAnalysis(
  Analyte = list(genes = significant_genes),
  universe = list(genes = measured_panel_genes),
  database = database, gene_index = index,
  min_path_size = 5, pval_cutoff = NULL
)
attr(result, "enrichment")$gene_mapping$inputs
attr(result, "enrichment")$gene_mapping$groups
```

Foreground and universe use the same complete identity index. All RaMP records
of a resolved gene contribute to its union of pathway memberships, but the gene
counts once in K, M, U and overlap. The universe is fixed before size filtering;
zero-overlap eligible pathways remain in the BH family. `pval_cutoff` filters
FDR after correction. Missing identifiers are excluded with an explicit audit;
do not set U to the raw panel length to hide mapping failures.

Official SpaMTPdb workflows harmonize human genes automatically. A custom
bundle can provide `gene_reference`, or an index can be supplied explicitly.
Custom bundles without a reference retain RaMP-only behavior without a network
request. `gene_mapping = "ramp"` explicitly reproduces historical mapping or
uses a separately curated species-specific database; it is not an orthology
conversion switch. Previously saved gene-ID mappings and p-values may change
after identity reconciliation even though Fisher arithmetic is unchanged.

Use the same `gene_index` in `findRegionalPathways()` and
`pathwayNetworkPlots()` to keep annotation identity and provenance consistent.

## One membership index for enrichment, expression and plots

`buildPathwayIndex()` joins the gene identity index to RaMP memberships once.
A pathway's `database_size` counts unique identities after reconciliation and
conflict exclusion; `database_raw_size` counts original RaMP records.
`measured_size` and `coverage_fraction` describe the selected assay or measured
universe. `used_size` and `used_members` record what actually contributes to a
score or test. In regional analysis these can be smaller than the measured set
because of the requested ranking filters. A pathway score is an expression
summary, not evidence that the whole pathway is activated.

This small example runs offline. Its fictional gene has two RaMP records with
complementary pathway memberships; both contribute while the gene counts once.

```{r shared-membership-example}
reference_example <- data.frame(
  hgnc_id = c("HGNC:1", "HGNC:2", "HGNC:3"),
  symbol = c("GENEA", "GENEB", "GENEC"), entrez_id = c("101", "102", "103"),
  uniprot_ids = c("U1", "U2", "U3")
)
database_example <- list(
  source_df = data.frame(rampId = paste0("RAMP_G_", c(1, 2, 3, 4)),
    sourceId = c("entrez:101", "uniprot:U1", "entrez:102", "entrez:103")),
  analytehaspathway = data.frame(rampId = paste0("RAMP_G_", c(1, 3, 2, 4)),
    pathwayRampId = c("P1", "P1", "P2", "P2")),
  pathway = data.frame(pathwayRampId = c("P1", "P2"), pathwayName = c("First", "Second")),
  gene_reference = reference_example
)
pathway_index_example <- buildPathwayIndex(database_example)
example <- SpaMTPData::spaMTPExampleData()
E_example <- matrix(seq_len(36), 3, dimnames = list(c("GENEA", "GENEB", "GENEC"), colnames(example)))
SingleCellExperiment::altExp(example, "transcriptome") <-
  SingleCellExperiment::SingleCellExperiment(assays = list(logcounts = log1p(E_example)))
example <- createPathwayAssay(example, analyte_type = "genes", assay = "transcriptome",
  slot = "logcounts", new_assay = "gene_ids", pathway_index = pathway_index_example)
example <- createPathwayObject(example, assay = "gene_ids", new.assay = "gene_scores",
  pathway_index = pathway_index_example)
SummarizedExperiment::rowData(SingleCellExperiment::altExp(example, "gene_scores"))[
  , c("pathwayRampId", "database_size", "measured_size", "coverage_fraction", "used_members")]
```

For the real human experiment loaded above, use the full measured RNA panel
and the same index for every entry point. This resource-dependent chunk is not
run during package builds; the real-data validation uses all 339 RNA features
and all 37,886 observations of human_brain_integrated 1.1.0.

```{r human-pathway-closure, eval=FALSE}
pathway_index <- buildPathwayIndex(database = database, gene_index = index)
spe <- createPathwayAssay(spe, analyte_type = "genes", assay = "SPT",
  slot = "counts", new_assay = "gene_ids", pathway_index = pathway_index)
spe <- createPathwayObject(spe, assay = "gene_ids", new.assay = "gene_pathways",
  pathway_index = pathway_index, min_path_size = 5, max_path_size = 500)
pathways <- SingleCellExperiment::altExp(spe, "gene_pathways")
coverage <- S4Vectors::metadata(pathways)$pathway_mapping$coverage
coverage[, c("pathwayRampId", "database_size", "database_raw_size",
             "measured_size", "coverage_fraction", "excluded_conflict_count")]
coverage$used_members
coverage$excluded_conflict_records
selected <- head(coverage$pathwayRampId[coverage$eligible], 2)
plots <- plotPathwaysSpatially(selected, spe, assay = "SPT", slot = "counts",
  pathway_index = pathway_index)
attr(plots, "pathway_coverage")
```

The original assays, feature names, coordinates and paired modalities are
preserved. A derived gene assay keeps its selected layer name: logcounts do
not become counts. Different expression rows mapping to one gene error by
default; choose `duplicate_genes = "mean"` or `"sum"` only after considering
the measurement scale (summing log-expression is generally inappropriate).
Identical repeated rows are deduplicated. Named pathway plots accept exact
`pathwayRampId` values; ambiguous display names error instead of choosing the
first match. Gene and pathway reference fingerprints are retained in outputs.

Fisher's complete coverage table is `attr(result, "pathway_coverage")`; its
per-row size columns also appear in the result. Pathway score objects put
coverage in rowData and the complete audit in metadata. Named plots return
coverage attributes and the exact plotted score vector. Unmeasured and
conflict-only pathways stay in the complete audit even when no score is drawn.
