---
title: "SpaMTP: Modern Data Access and Workflow Migration"
output:
  BiocStyle::html_document:
    toc: true
vignette: >
  %\VignetteIndexEntry{SpaMTP: Modern Data Access and Workflow Migration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "##", fig.align = "center")
```

# One container, public accessors

The submission-branch API uses existing Bioconductor S4 containers. Raw,
unaligned spectra remain in Cardinal; a common feature-by-pixel matrix belongs
in `SpatialExperiment`. No knowledge of internal S4 slots is needed to read,
modify or subset these objects. The historical Seurat tutorials remain on the
published-workflow branch; their code is not the submission-branch API.

| Data to access | Public interface |
|---|---|
| Expression/intensity matrices | `assay()`, `assayNames()` |
| Feature m/z and annotations | `rowData()` |
| Pixel metadata and groups | `colData()`, `colLabels()` |
| Pixel coordinates | `spatialCoords()`, `spatialCoordsNames()` |
| Optical images | `imgData()`, `imgRaster()`, `addImg()` |
| Paired modalities | `altExp()`, `altExpNames()` |
| PCA and other embeddings | `reducedDim()`, `reducedDimNames()` |
| Analysis provenance | `metadata()` |

These accessors also have replacement forms. For example,
`colData(spe)$region <- values` modifies pixel metadata through the container's
public API. A `slot = "counts"` argument retained by some SpaMTP functions is a
compatibility spelling for the expression assay name; it is not an instruction
to inspect an S4 slot.

```{r load-package, message=FALSE}
library(SpaMTP)
```

# Read a binned MSI table

The CSV contains one row per pixel, `x` and `y` coordinate columns, and one
column per measured m/z. This runnable example writes a small temporary input
file. Replace it with your own path for real data. Raw imzML processing is
demonstrated in the [native workflow](SpaMTP_Bioconductor.html).

```{r read-matrix}
input <- data.frame(
  x = rep(1:3, each = 2), y = rep(1:2, 3),
  check.names = FALSE,
  "100.01" = c(0, 2, 3, 5, 7, 9),
  "200.02" = c(4, 1, 6, 2, 8, 3),
  "300.03" = c(2, 3, 1, 4, 2, 6)
)
spe <- local({
  path <- tempfile(fileext = ".csv")
  on.exit(unlink(path))
  utils::write.csv(input, path, row.names = FALSE)
  readSMMatrix(path, project.name = "section1", verbose = FALSE)
})
stopifnot(methods::is(spe, "SpatialExperiment"), nrow(spe) == 3, ncol(spe) == 6)
SummarizedExperiment::assayNames(spe)
SummarizedExperiment::assay(spe, "counts")
SummarizedExperiment::rowData(spe)$mz
```

# Pixel and feature metadata

Use feature identifiers to index matrix rows, and pixel identifiers to index
columns. m/z is explicit feature metadata, not something to recover by editing
feature names. For discrete plotting groups, use a factor even if labels were
originally stored as integer cluster numbers.

```{r metadata-access}
SummarizedExperiment::colData(spe)$region <- factor(rep(c("edge", "core"), 3))
SingleCellExperiment::colLabels(spe) <- spe$region
SummarizedExperiment::rowData(spe)$feature_label <- c("peak A", "peak B", "peak C")

intensity <- SummarizedExperiment::assay(spe, "counts")
SummarizedExperiment::colData(spe)$total_intensity <- Matrix::colSums(intensity)
SummarizedExperiment::colData(spe)$detected_features <- Matrix::colSums(intensity > 0)
as.data.frame(SummarizedExperiment::colData(spe))
```

QC values above are summaries of MSI intensities, not UMI counts. They are
computed from the selected matrix instead of assuming that historical
Seurat-generated QC columns still describe the current features and assay.

```{r expression-access, fig.width=6, fig.height=4}
spe <- normalizeSMData(spe, "LogNormalize", verbose = FALSE)
SummarizedExperiment::assayNames(spe)
scater::plotExpression(
  spe, features = rownames(spe)[1], x = "region", exprs_values = "logcounts"
)
```

# Coordinates and optical images

`spatialCoords()` is the coordinate source for native spatial analysis.
Historical `x_coord`/`y_coord` metadata may describe an earlier alignment;
retaining these columns for provenance does not make them authoritative.
SpaMTP expects named `x` and `y` axes in a shared coordinate system. If an
importer uses names such as `pxl_col_in_fullres` and `pxl_row_in_fullres`, map
the known column axis to `x` and row axis to `y` with `spatialCoordsNames()`.
Renaming does not convert units, reverse an axis or perform image registration.

```{r coordinate-access}
xy <- SpatialExperiment::spatialCoords(spe)
SpatialExperiment::spatialCoordsNames(spe)
shifted <- spe
SpatialExperiment::spatialCoords(shifted)[, "x"] <- xy[, "x"] + 10
stopifnot(identical(SpatialExperiment::spatialCoords(spe), xy))

# Demonstrate an explicit name mapping on a copy, without changing the values.
imported <- spe
SpatialExperiment::spatialCoordsNames(imported) <-
  c("pxl_col_in_fullres", "pxl_row_in_fullres")
axisNames <- SpatialExperiment::spatialCoordsNames(imported)
axisNames[match(c("pxl_col_in_fullres", "pxl_row_in_fullres"), axisNames)] <- c("x", "y")
SpatialExperiment::spatialCoordsNames(imported) <- axisNames
stopifnot(identical(SpatialExperiment::spatialCoords(imported), xy))
```

An image has its own sample ID, image ID and scale factor in `imgData()`.
Adding a raster does not register it to the MSI pixels. This small artificial
image illustrates storage only. With `load = TRUE`, the raster remains usable
after the temporary PNG is removed; a path-backed image requires a lasting path.

```{r image-access}
spe <- local({
  path <- tempfile(fileext = ".png")
  on.exit(unlink(path))
  png::writePNG(matrix(seq(0, 1, length.out = 12), nrow = 3), path)
  addSpatialImage(spe, imageSource = path, sampleId = "section1",
    imageId = "demo", scaleFactor = 1, load = TRUE)
})
SpatialExperiment::imgData(spe)[, c("sample_id", "image_id", "scaleFactor")]
raster <- SpatialExperiment::imgRaster(spe, sample_id = "section1", image_id = "demo")
stopifnot(inherits(raster, "raster"))
```

# Paired modalities and standard subsetting

Only measurements of the same pixels belong in an `altExp()`. First align and
map independent MSI pixels and transcriptomic spots with `mapSpatialOmics()`;
do not rename unrelated spots to force a match. `addTranscriptome()` validates
pixel names and reorders an already paired matrix.

```{r paired-access}
transcriptome <- matrix(seq_len(12), nrow = 2,
  dimnames = list(c("gene1", "gene2"), rev(colnames(spe))))
spe <- addTranscriptome(spe, transcriptome)
rna <- SingleCellExperiment::altExp(spe, "transcriptome")
stopifnot(identical(colnames(rna), colnames(spe)))
SummarizedExperiment::assay(rna, "counts")

selected <- spe[, spe$region == "edge"]
stopifnot(identical(colnames(SingleCellExperiment::altExp(selected, "transcriptome")),
  colnames(selected)))
stopifnot(nrow(SpatialExperiment::spatialCoords(selected)) == ncol(selected))
```

Feature subsetting affects the primary experiment's rows; each alternative
experiment retains its own feature space. Standard column subsetting keeps
coordinates and alternative experiments aligned to the selected pixels.

```{r provenance-access}
S4Vectors::metadata(spe)$analysis <- list(
  input_type = "binned MSI", normalization = "LogNormalize")
S4Vectors::metadata(spe)$analysis
```

For full annotation examples, see the
[indexed annotation pipeline](Metabolite_Annotation_Pipeline.html).
It reads candidate summaries from `rowData()` and stores ranked candidates
and annotation provenance in `metadata()`.

# Optional Seurat boundary

SeuratObject is optional and is needed only for explicit conversion. In Seurat
5, `LayerData()` reads an exact expression layer, double brackets read cell or
feature metadata, and `GetTissueCoordinates()` reads image/FOV coordinates.
SpaMTP's converters use these public APIs. Older serialized objects that cannot
be read by current accessors should be updated explicitly with
`SeuratObject::UpdateSeuratObject()` in the external Seurat workflow; SpaMTP
does not patch their internal slots.

```{r optional-conversion}
if (requireNamespace("SeuratObject", quietly = TRUE)) {
  seurat <- spatialExperimentToSeurat(spe)
  values <- SeuratObject::LayerData(seurat, assay = "Spatial", layer = "counts")
  pixel_metadata <- seurat[[]]
  feature_metadata <- seurat[["Spatial"]][[]]
  centres <- SeuratObject::GetTissueCoordinates(seurat, image = "fov", full = FALSE)

  converted <- seuratToSpatialExperiment(seurat, image = "fov", layer = "counts")
  stopifnot(identical(colnames(converted), colnames(spe)))
  stopifnot(isTRUE(all.equal(
    as.matrix(SummarizedExperiment::assay(converted, "counts")), as.matrix(values))))
  S4Vectors::metadata(converted)$seurat_interoperability
}
```

When no image is specified, a single image/FOV is authoritative over any
historical metadata coordinates. Metadata is used automatically only if there
are no images. With multiple images, select `image` or supply explicit named
`coordinates`; the converter never guesses which slice to use. Centroids are
read as one centre per pixel, not as expanded boundary vertices.

Only the requested exact layer is transferred. Join split Seurat layers first
if the complete dataset is intended. Unpaired assays are skipped with a warning.
Optical rasters, graphs and reductions are not transferred. After conversion,
continue with native accessors; Seurat is not used for SpaMTP analysis.

```{r session-info}
sessionInfo()
```
