---
title: "SpaMTP: Spatial Metabolite Annotation"
author: "Tianyao Lu"
output:
  BiocStyle::html_document:
    toc: true
vignette: >
  %\VignetteIndexEntry{SpaMTP: Spatial Metabolite Annotation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "##",
  fig.align = "center"
)
```

# Overview

`SpaMTP` provides analysis and visualisation methods for spatial metabolomics
and paired spatial multi-omics experiments. Its primary containers are
Cardinal's `MSImagingArrays`/`MSImagingExperiment` and Bioconductor's `SpatialExperiment`.
Seurat is not an analysis dependency. Only the optional boundary-conversion
example uses SeuratObject; all analysis runs on native containers. Explicit
conversion functions connect existing Seurat workflows. Versioned
annotation resources are supplied by the companion
`SpaMTPdb` package and experiment data by `SpaMTPData`.

This vignette demonstrates a small, fully reproducible accurate-mass annotation
workflow. It deliberately uses an in-memory database so that package checking
does not require a network connection or a large data download.

See [modern data access](Modern_Data_Access.html) for replacing historical
container-internal reads and writes, and the
[indexed annotation pipeline](Metabolite_Annotation_Pipeline.html) for the
complete native annotation tutorial.
The [companion workflow](SpaMTP_Companion_Workflow.html) demonstrates how
SpaMTPData, SpaMTPdb and SpaMTP work together without downloads or Seurat.

```{r load-package, message=FALSE}
library(SpaMTP)
```

# Bioconductor container model

The container depends on the data's processing stage:

| Processing stage | Container | SpaMTP mapping |
|---|---|---|
| Raw or continuous spectra | Cardinal `MSImagingArrays` / `MSImagingExperiment` | File-backed spectra and unequal mass axes remain in Cardinal/matter |
| Aligned or binned MSI | `SpatialExperiment` | Intensities in `assays()`, m/z and annotations in `rowData()`, pixel metadata in `colData()` |
| Pixel coordinates | `SpatialExperiment` | `spatialCoords()` |
| Optical images | `SpatialExperiment` | `imgData()` |
| Paired transcriptome | `SpatialExperiment` | `altExp()` |

Raw imzML should be read and processed in Cardinal without materialising the
spectra. Once features share a common axis, conversion is explicit:

```{r cardinal-workflow, eval=FALSE}
raw <- Cardinal::readImzML("section.imzML", memory = FALSE, as = "MSImagingArrays")
raw <- normalizeSMData(raw) # Cardinal queues TIC processing.
spe <- binSpaMTP(raw, resolution = 0.01, units = "mz")
```

The S4 binning method executes Cardinal's queued processing and obtains a
common axis before creating an SPE. The raw-array mass range is scanned across
all spectra, not inferred from the first pixel alone. Calling
`asSpatialExperiment()` on unaligned `MSImagingArrays` gives an explicit request
to bin first; no resolution is guessed during conversion.

The reverse coercion is also registered, so either the SpaMTP generic or the
standard S4 coercion can be used:

```{r cardinal-coercion, eval=FALSE}
cardinal <- asCardinal(spe)
cardinal <- methods::as(spe, "MSImagingExperiment")
```

# A binned MSI workflow

The following small object illustrates the native layout without external
files.

```{r spatial-experiment}
intensity <- matrix(
  seq_len(24),
  nrow = 4,
  dimnames = list(
    paste0("mz-", c(100.01, 100.02, 101.01, 101.02)),
    paste0("pixel", seq_len(6))
  )
)
coordinates <- cbind(
  x = rep(seq_len(3), each = 2),
  y = rep(seq_len(2), times = 3)
)
spe <- SpatialExperiment::SpatialExperiment(
  assays = list(counts = intensity),
  rowData = S4Vectors::DataFrame(
    mz = c(100.01, 100.02, 101.01, 101.02)
  ),
  colData = S4Vectors::DataFrame(
    region = rep(c("edge", "core"), 3)
  ),
  spatialCoords = coordinates,
  sample_id = rep("sample01", ncol(intensity))
)

spe <- normalizeSMData(spe, normalisation.type = "LogNormalize")
plotSpatialFeature(spe, features = "mz-100.01", assayName = "logcounts")
```

For paired measurements, transcriptomic data are stored as an alternative
experiment and matched by pixel name. Integration uses Bioconductor reduced
dimensions rather than Seurat assay internals.

```{r alternative-experiment}
transcriptome_sce <- SingleCellExperiment::SingleCellExperiment(
  assays = list(counts = matrix(
    c(1, 2, 3, 4, 4, 2, 1, 3, 2, 5, 4, 1,
      3, 1, 5, 2, 1, 4, 2, 5, 5, 3, 1, 4),
    nrow = 4,
    dimnames = list(paste0("gene", seq_len(4)), colnames(spe))
  ))
)
spe <- addTranscriptome(spe, transcriptome_sce, name = "transcriptome")
spe <- normalizeSMData(spe, "LogNormalize", assay = "transcriptome")
set.seed(42)
spe <- multiOmicIntegration(spe, dims.list = list(1, 1))
SingleCellExperiment::altExpNames(spe)
SingleCellExperiment::reducedDim(spe, "integrated")

# Reuse scater directly for standard expression visualisation.
scater::plotExpression(
  spe,
  features = "mz-100.01",
  x = "region",
  exprs_values = "logcounts"
)
```

The joint embedding concatenates standardized PCA scores with equal weight
per modality. It is **not Seurat WNN**, and does not claim equivalent results.
Only `main` and `transcriptome` are included by default; supply `modalities`
explicitly to include another measured modality. Derived pathway scores are
not included automatically.

## Alignment and pixel mapping

Spatial mapping uses matching `sample_id` values and a common coordinate
system. Independent slices are never connected simply because their x/y
coordinates overlap. The example below maps MSI to paired transcriptomic
pixel centres; lower-resolution spots require an explicit radius or polygon.

```{r native-mapping}
target <- SpatialExperiment::SpatialExperiment(
  assays = SummarizedExperiment::assays(transcriptome_sce),
  colData = SummarizedExperiment::colData(spe),
  spatialCoords = SpatialExperiment::spatialCoords(spe)
)
mapped <- mapSpatialOmics(
  spe, target, ST.hires = TRUE, SM.pixel.width = 1
)
stopifnot(all(mapped$mapped), all(mapped$n_source_pixels == 1))
SingleCellExperiment::altExpNames(mapped)

affine <- diag(3)
affine[1:2, 3] <- c(0.2, -0.1)
aligned <- alignSpatialOmics(spe, target, transformation = affine)
checkAlignment(aligned, target)
```

MSI values are averaged over matched source pixels. Unmapped targets are
retained with `mapped = FALSE` and zero-filled intensities unless
`dropUnmapped = TRUE`. These missing measurements must not be interpreted as
observed biological zeros. Optical images are stored with `addSpatialImage()`
and `imgData()`; their scale factors change display dimensions, not the stored
spatial coordinates.

## Binning and statistical reuse

```{r native-binning}
binned <- binSpaMTP(spe, resolution = 0.05, units = "mz")
stopifnot(isTRUE(all.equal(
  as.numeric(Matrix::colSums(SummarizedExperiment::assay(binned, "counts"))),
  as.numeric(Matrix::colSums(SummarizedExperiment::assay(spe, "counts")))
)))
```

Aligned-matrix binning assigns each feature once and conserves total intensity
under sum aggregation. Raw Cardinal input retains Cardinal's reference-window
binning semantics. Re-annotate the newly binned features before using
annotation-dependent analyses.

TMM factors are computed by `edgeR::normLibSizes()`. Differential testing uses
`limma::lmFit()` and `limma::treat()`, while `runPooling()` uses
`edgeR::sumTechReps()` for technical partitions. Random pixel pools are not
independent biological replicates and cannot support population-level inference
as if they were. Use independent biological samples for such inference.

# Native feature plots and annotation ranking

MSI plots read masses and annotations from `rowData()`, including when feature
IDs are not formatted as `mz-...`. A `plusminus` window includes every peak
within the tolerance of the nearest measured mass. Annotation plots sum all
matching peaks; overlapping windows count each peak once.

```{r native-feature-plots}
SummarizedExperiment::rowData(spe)$all_IsomerNames <-
  c("Candidate A", "Candidate A", "Candidate B", "Candidate C")
imageMZAnnotationPlot(spe, "Candidate A", slot = "counts",
                      dark.background = FALSE)
massIntensityPlot(spe, group.by = "region", mz.labels = 100.01)
```

Use `spatialMZPlot(spe, 100.01, images="optical")` after attaching an image with
`addSpatialImage()`. Its `imgData()` scale factor relates raster pixels to
`spatialCoords()` units. Samples are displayed separately. Former Seurat
segmentation/molecule/blending arguments are not supported. Interactive mass
windows are available without Seurat; creating the app does not launch it:

```{r native-window-app}
app <- interactiveSpatialPlot(spe)
class(app)
# shiny::runApp(app)
```

The following synthetic example ranks three candidates for one feature using
precomputed paired pathway expression. Real workflows can create these scores
with `createPathwayObject()`, or let `calculateAnnotationStatistics()` reuse it
on an alternative experiment indexed by RaMP analyte IDs.

```{r native-annotation-ranking}
ranking_demo <- spe
SummarizedExperiment::rowData(ranking_demo)$all_Ramp_IDs <-
  rep("RAMP_C_1; RAMP_C_2; RAMP_C_3", nrow(ranking_demo))
pathway_values <- rbind(pA = 1:6, pB = c(2, 1, 5, 3, 6, 4), pC = 6:1)
colnames(pathway_values) <- colnames(ranking_demo)
SingleCellExperiment::altExp(ranking_demo, "pathway") <-
  SingleCellExperiment::SingleCellExperiment(
    assays = list(pathwayScores = pathway_values))
rank_db <- list(
  source_df = data.frame(rampId = paste0("RAMP_C_", 1:3),
                         commonName = c("A", "B", "C")),
  analytehaspathway = data.frame(
    rampId = c("RAMP_C_1", "RAMP_C_1", "RAMP_C_2", "RAMP_C_3"),
    pathwayRampId = c("pA", "pB", "pB", "pC")))
calculateSingleAnnotationStatistics(100.01, ranking_demo, database = rank_db)
```

This retains the historical weighted z-score formula, now calculated directly
from paired Pearson correlations. Its standard-normal tail and BH-adjusted
values are **heuristic ranking measures, not calibrated identification
p-values**. Correlated pathways and spatially dependent pixels violate the
simple interpretation of that tail probability. Constant expression has
undefined correlation; a constant candidate-score component contributes zero.

# Optional Seurat conversion boundary

SpaMTP analysis does not load Seurat or SeuratObject. Only these explicit
converters need SeuratObject. Convert existing data once, then continue with
native normalization, scaling, reductions and plotting.

```{r optional-seurat}
if (requireNamespace("SeuratObject", quietly = TRUE)) {
    seurat_object <- spatialExperimentToSeurat(spe, includeAltExps = FALSE)
    spe_again <- seuratToSpatialExperiment(seurat_object)
    sce_again <- seuratToSingleCellExperiment(seurat_object)
    stopifnot(identical(colnames(spe_again), colnames(sce_again)))
}
```

Use seuratToSingleCellExperiment() for non-spatial transcriptomics; it never
invents spatial coordinates. For spatial input, coordinates must have x/y
columns and named pixels, or be supplied by a suitable Seurat image. Multiple
images require an explicit selection or coordinate table. A single image/FOV
is read through its public spatial accessor and takes precedence over cached
metadata coordinates, which may predate alignment. The selected source is
recorded in `metadata(spe_again)$seurat_interoperability`. Only one exact
layer is transferred; split layers are not silently partially imported.
Optical rasters, reductions and graphs are not transferred. Paired alternative
assays are matched by pixel IDs, and unpaired assays are skipped with a warning.

For an existing Seurat MSI workflow, the native sequence is:

```{r native-scaling}
scaled_spe <- normalizeSMData(spe, "LogNormalize", verbose = FALSE)
scaled_spe <- scaleSMData(scaled_spe)
SummarizedExperiment::assayNames(scaled_spe)
```

scaleSMData() centres/scales each feature across pixels and stores a separate
scaled assay. This creates a dense matrix. Unlike the historical WNN workflow,
multiOmicIntegration() combines equal-weight PCA embeddings; the algorithms
must not be described as numerically equivalent.

# Inspect available database resources

`spaMTPDatabaseInfo()` reads the lightweight SpaMTPdb registry without
downloading any annotation table.

```{r database-registry}
database_registry <- spaMTPDatabaseInfo(version = "latest")
database_registry[
  database_registry$resource %in% c("chem_props", "smiles_features"),
  c("resource", "version", "category", "rows", "columns")
]
```

# Create a small structure-aware database

The example contains D-2-hydroxyglutaric acid and glucose. SMILES connectivity
is decomposed into interpretable functional groups and ionisation sites before
candidate generation.

```{r demo-database}
demo_database <- data.frame(
  ramp_id = c("RAMP_C_2HG", "RAMP_C_GLUCOSE"),
  chem_source_id = c("HMDB:HMDB0000606", "HMDB:HMDB0000122"),
  common_name = c("D-2-hydroxyglutaric acid", "D-glucose"),
  monoisotop_mass = c(148.037173, 180.063388),
  mol_formula = c("C5H8O5", "C6H12O6"),
  iso_smiles = c("O=C(O)CC(O)C(=O)O", "OCC1OC(O)C(O)C(O)C1O"),
  stringsAsFactors = FALSE
)

demo_database <- annotateSMILESStructure(demo_database)
demo_database[, c(
  "common_name", "carboxyl_sites", "hydroxyl_sites",
  "max_exchangeable_protons", "positive_mode_score",
  "negative_mode_score", "alkali_affinity_score"
)]
```

# Build and query an indexed search space

Candidate masses are generated once and sorted. Repeated observed peaks can
then be queried using a binary range search rather than scanning the full
database for every peak.

```{r indexed-search}
annotation_index <- buildMZAnnotationIndex(
  db = demo_database,
  polarity = "negative",
  adducts = c("[M-H]-", "[2M-H]-"),
  infer_structure = "always"
)

observed_mz <- 147.02935
candidates <- queryMZAnnotationIndex(
  observed_mz = observed_mz,
  index = annotation_index,
  ppm = 5
)

candidates[, c(
  "observed_mz", "expected_mz", "metabolite_names", "adduct",
  "ppm_error", "score"
)]
```

# Reproducibility

For production analyses, record the database version returned by
`spaMTPDatabaseInfo()` together with ion mode, matrix profile, adduct rules,
mass tolerance, and annotation-score threshold.

```{r session-info}
sessionInfo()
```
