---
title: "SpaMTP, SpaMTPdb and SpaMTPData: A Native Companion Workflow"
author: "Tianyao Lu"
output:
  BiocStyle::html_document:
    toc: true
vignette: >
  %\VignetteIndexEntry{SpaMTP, SpaMTPdb and SpaMTPData: A Native Companion Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "##", fig.align = "center")
seuratBefore <- intersect(c("Seurat", "SeuratObject"), loadedNamespaces())
```

# Clear responsibilities

| Package | Responsibility | Primary representations |
|---|---|---|
| SpaMTPData | Discoverable experiment resources and a small synthetic example | Cardinal, SpatialExperiment, auxiliary files; labelled historical archives |
| SpaMTPdb | Versioned annotation and pathway resources | Independent tables/lists with checksums and provenance |
| SpaMTP | Analysis, annotation, plotting and explicit conversion | Existing Bioconductor S4 containers and public accessors |

The data packages do not import SpaMTP or Seurat. SpaMTP imports SpaMTPdb and
suggests SpaMTPData for demonstrations; there is no circular runtime dependency.
Use SpaMTP >= 0.99.4, SpaMTPdb >= 0.99.3 and SpaMTPData >= 0.99.4 together.
The package versions are not the same thing as RaMP or experiment-data versions.

All evaluated examples below work without downloads. The experiment and small
chemical table are explicitly synthetic teaching inputs, not a biological
dataset or a replacement for the complete RaMP snapshot.

```{r packages, message=FALSE}
library(SpaMTP)
stopifnot(requireNamespace("SpaMTPData", quietly = TRUE))
```

# Data to a native analysis container

```{r native-data}
spe <- SpaMTPData::spaMTPExampleData()
stopifnot(methods::is(spe, "SpatialExperiment"))
SummarizedExperiment::assayNames(spe)
SummarizedExperiment::rowData(spe)$mz
SingleCellExperiment::altExpNames(spe)
```

Counts/intensities, feature metadata, pixel metadata and coordinates have
separate public accessors. The paired transcriptome already has matching pixel
names and is stored in `altExp()`. Independent measurements must first be
spatially aligned and mapped; they cannot be paired merely by renaming columns.

```{r preprocess, fig.width=6, fig.height=4}
spe <- normalizeSMData(spe, "LogNormalize", verbose = FALSE)
spe <- runMetabolicPCA(spe, npcs = 2, slot = "logcounts")
plotSpatialFeature(spe, features = "peak1", assayName = "logcounts",
  labels = "Synthetic peak 1", pointSize = 4)
SingleCellExperiment::reducedDimNames(spe)
```

# Database discovery without network access

The registry is lightweight package metadata. Both access paths below describe
the same resource release and use the coordinated camelCase API.

```{r database-registry}
registry <- SpaMTPdb::spaMTPdbResources(version = "3.0.7")
stopifnot(identical(registry, spaMTPDatabaseInfo(version = "3.0.7")))
registry[registry$resource %in% c("chem_props", "smiles_features", "pathway"),
  c("resource", "version", "r_data_class", "bytes", "md5")]
```

For a production analysis, `loadSpaMTPDatabase()` uses the same resource reader
and retains database provenance. A staged official file must match its size and
checksum; `offline = TRUE` can also reuse a verified download cache.

```{r production-resources, eval=FALSE}
database <- loadSpaMTPDatabase(
  c("chem_props", "pathway"), version = "3.0.7",
  local_dir = "/path/to/SpaMTPdb-resources/3.0.7", offline = TRUE
)
```

# Annotate without inventing a downloaded database

This small custom table contains two neutral masses and explicit illustrative
identifiers. It is supplied as a custom database rather than being placed under
an official resource filename with an incorrect checksum.

```{r annotation}
custom <- data.frame(
  id = c("demo-2HG", "demo-glucose"),
  name = c("2-hydroxyglutarate", "glucose"),
  formula = c("C5H8O5", "C6H12O6"),
  exactmass = c(148.037173366, 180.063388104)
)
index <- buildMZAnnotationIndex(custom, adducts = "M+H", polarity = "positive")
annotated <- annotateSM(spe, index = index, adducts = "M+H",
  ppm_error = 5, return.only.annotated = FALSE, verbose = FALSE)
stopifnot(identical(dim(annotated), dim(spe)),
  identical(colnames(SingleCellExperiment::altExp(annotated, "transcriptome")),
    colnames(spe)))
SummarizedExperiment::rowData(annotated)[,
  c("mz", "all_IsomerNames", "all_Scores")]
annotationInfo(annotated)
```

These accurate-mass candidates are hypotheses, not confirmed identifications
or calibrated probabilities. The [full annotation tutorial](Metabolite_Annotation_Pipeline.html)
explains index reuse, contextual scoring and interpretation.

# Published native datasets and historical archives

The published experiment registry records what the source files actually
contain. The default resource release, `1.1.0`, supplies seven native
SpatialExperiment datasets with new URLs and hashes in
[Zenodo record 22733262](https://zenodo.org/records/22733262).
Historical Seurat files remain explicitly available in release `1.0.0`.

```{r inspect-native}
native_metadata <- SpaMTPData::spaMTPData("mouse_brain_fmp10", metadata = TRUE)
stopifnot(native_metadata$version == "1.1.0",
  native_metadata$r_data_class == "SpatialExperiment",
  !native_metadata$requires_conversion)
```

```{r inspect-archive}
archive <- SpaMTPData::spaMTPData("mouse_brain_fmp10",
  version = "1.0.0", metadata = TRUE)
archive[, c("resource", "version", "r_data_class", "requires_conversion")]
stopifnot(identical(intersect(c("Seurat", "SeuratObject"), loadedNamespaces()),
  seuratBefore))
```

The one-time preparation recipe shipped by SpaMTPData creates a separate
native RDS for each of the seven historical archives. The published files can
now be retrieved directly; no Seurat conversion is needed when using them:

```{r read-published-native, eval=FALSE}
native <- SpaMTPData::spaMTPData("mouse_brain_dhb_striatum", version = "1.1.0")
native <- normalizeSMData(native, verbose = FALSE)
SingleCellExperiment::altExpNames(native) # SPT
```

The native files have their own checksum/provenance manifest and do not
replace the original archive URLs or checksums. They preserve
counts and any complete `data` layer; the latter is not automatically relabelled
as `logcounts`. Optical rasters, polygons and Seurat graphs/weights/reductions
are omitted. Source annotations, including historical `db_3`, retain their
original interpretation and are not upgraded to new scored annotations.

To work with a historical archive instead, conversion remains explicit:

```{r convert-archive, eval=FALSE}
legacy <- SpaMTPData::spaMTPData("mouse_brain_fmp10",
  version = "1.0.0", offline = TRUE)
native <- seuratToSpatialExperiment(legacy, assay = "Spatial", image = "fov")
```

Select coordinates, layers and images deliberately. Optical rasters, graphs
and reductions are not transferred. See [modern data access](Modern_Data_Access.html)
for the conversion contract and [the native workflow](SpaMTP_Bioconductor.html)
for further analysis. Older PascalCase resource calls must be updated along
with the coordinated companion package versions.

```{r session-info}
sessionInfo()
```
